---
title: "Emission Reduction from Renewable Energy in Virginia and North Carolina"
author: "Jonathan M. Gilligan"
date: "12/4/2019"
output: 
  pdf_document: default
---

```{r knitr, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache=FALSE)
```
```{r setup, echo=FALSE, eval=TRUE, include=FALSE, message=FALSE, warning=FALSE}
```
```{r dl_egrid_data, echo=FALSE, eval=TRUE, include=FALSE, message=FALSE, warning=FALSE}
```
```{r load_data, echo=FALSE, eval=TRUE, include=FALSE, message=FALSE, warning=FALSE}
```
```{r emissions_factor, echo=FALSE, eval=TRUE, include=FALSE, message=FALSE, warning=FALSE}
```
# Introduction

High tech companies have stimulated the adoption of 750 MW nameplate capacity of new renewable electricity
generation capacity in North Carolina and Virginia. I estimate the reduction in annual greenhouse gas 
emissions from using renewables instead of building an equivalent amount of natural gas generation.

The result of my calculations is a reduction of annual emissions by `r scales::comma(gas_emissions, accuracy = 1000)` metric 
tons of CO~2~e

# Methods

```{r setup, echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE}
library(pacman)
p_load(tidyverse, magrittr, lubridate, readxl, broom, janitor)

data_dir = here::here("data")
```

First, we download the data file from EPA (2018) with generation, capacity
factors, and emissions factors for each generating plant in each state.

```{r dl_egrid_data, echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE}
egrid_file <- file.path(data_dir, "egrid2016_data.xlsx")
egrid_url <- "https://www.epa.gov/sites/production/files/2018-02/egrid2016_data.xlsx"
if (! file.exists(egrid_file)) download.file(egrid_url, egrid_file, mode = "wb")
```

Since almost all new fossil fuel generation capacity is natural
gas, assume that if the renewable generating capacity were not installed, an
equivalent amount of natural gas capacity would have to be added to supply the
data centers.

For our analysis, we break out the renewable and natural gas plants in Virginia
and North Carolina.

The table columns we use are:
* `pstatabb`: Two-letter state abbreviation.
* `plfuelct`: Fuel category (`SOLAR`, `WIND`, or `GAS` for our cases).
* `capfact`: Capacity factor for that generating plant
* `namepcap`: Nameplate generating capacity (MW) for that plant.
* `plngenan`: Annual generation for that plant (MWh).
* `plco2eqa`: Annual GHG emissions for the plant (short tons CO~2~e).
* `plc2erta`: Annual GHG emissions rate for the plant (pound CO~2~e per MWh).

```{r load_data, echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE}
egrid <- read_excel(egrid_file, "PLNT16",
                    skip = 1, na = c("", "NA", "N/A", "na", "n/a")) %>%
  clean_names()

renewables <- egrid %>%
  filter(pstatabb %in% c("NC", "VA"), plfuelct %in% c("SOLAR", "WIND")) %>%
  select(pstatabb, plfuelct, capfac, namepcap, plngenan, plco2eqa, plc2erta)

natgas <- egrid %>%
  filter(pstatabb %in% c("NC", "VA"), plfuelct %in% c("GAS")) %>%
  select(pstatabb, plfuelct, capfac, namepcap, plngenan, plco2eqa, plc2erta)
```

Now we calculate the equivalent amount of natural gas that would have to be
installed to replace renewables with a nameplate capacity of 750 MW.
We calculate the average capacity factor for each type of electricity
(the eGRID 2016 numbers apply to 2016, which was a leap year, so we use 
366 days for the length of the year).

The capacity factor is calculated as
$$ 
\text{capacity factor} = \frac{\sum_{p \in \text{plants}} \text{annual generation (MWh)}}{ \text{hours in a year} \times \sum_{p \in \text{plants}} \text{nameplate capacity (MW)}}.
$$
The capacity factors are:

* Renwwables `r scales::percent(renew_factor$cap_factor, accuracy = 0.1)`
* Natural gas: `r scales::percent(natgas_factor$cap_factor, accuracy = 0.1)`

The emissions factor is calculated as
$$
\text{emissions factor} = \frac{\sum_{p \in \text{plant}} \text{annual GHG emissions (tons)}}{\sum_{p \in \text{plant}} \text{annual generation (MWh)}},
$$
which gives an emissions factor for natural gas of `r scales::comma(natgas_factor$e_factor, accuracy = 0.001)` metric tons CO~2~e/MWh.

The generating capacity of natural gas that would be required to substitute for 750 MW nameplate capacity of renewables would be
$$
\text{cap}_{\text{gas}} = \text{cap}_{\text{renweables}} \times \frac{\text{cap factor}_{\text{renewables}}}{\text{cap factor}_{\text{gas}}}
$$
and the annual GHG emissions that would result would be 
$$
\begin{aligned}
\text{emissions} &= \text{emissions factor}_{\text{gas}} \times \text{cap}_{\text{gas}} \times \text{hours per year}\\
                 &= \text{emissions factor}_{\text{gas}} \times \text{cap}_{\text{renweables}} \times \frac{\text{cap factor}_{\text{renewables}}}{\text{cap factor}_{\text{gas}}} \times \text{hours per year}
\end{aligned}
$$

The results of the

```{r emissions_factor, echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE}
# 2016 is a leap year: use 366 day year.
renew_factor <- renewables %>% select(namepcap, plngenan) %>%
  na.omit() %>% summarize_all(sum) %>%
  mutate(cap_factor = plngenan / (24 * 366 * namepcap))

# emissions is given in short tons, so divide by 1.1 to get metric tons.
natgas_factor <- natgas %>% select(namepcap, plngenan, plco2eqa) %>%
  na.omit() %>% summarize_all(sum) %>%
  mutate(cap_factor = plngenan / (24 * 366 * namepcap),
         e_factor = plco2eqa / plngenan / 1.1)

# Metric tons CO2e
gas_emissions <- 750 * 24 * 365 * renew_factor$cap_factor * natgas_factor$e_factor

```

# References 

EPA 2018. _Emissions & Generation Resource Integrated Database (eGRID 2016)_, 
U.S. Environmental Protection Agency.
<https://www.epa.gov/energy/emissions-generation-resource-integrated-database-egrid>,
eGRID 2016 Data file:
`r str_c('<', egrid_url, '>')`.
Retrieved Nov. 26, 2019.

